---
title: "Foundations of Marketing Analytics: Module 1: Statistical Segmentation"
author: "Stefan Avey"
date: "2017-01-16"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    df_print: kable
---


```{r setup, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}

library(tidyverse)
library(lubridate)
library(aveytoolkit)

options(stringsAsFactors = FALSE)
```

# Load the Data

```{r loadData}

###############
## Load data ##
###############
purchases <- read.delim("data/purchases.txt", header = FALSE)

###################
## Preprocessing ##
###################
## Add column names
colnames(purchases) <- c("customer_id", "purchase_amount", "date_of_purchase")

## Convert date and add column for purchase year
purchases <- purchases %>%
    mutate(date_of_purchase = ymd(date_of_purchase)) %>%
    mutate(year_of_purchase = year(date_of_purchase))

## Look at the data
head(purchases)
summary(purchases)

```

# Compute Recency, Frequency, and Monetary Value

Three common characteristics of customers readily available from a transactional database are recency, frequency, and monetary value.

1. *Recency*: Time since last purchase.
2. *Frequency*: Number of purchases made in the past.
3. *Monetary value*: Amount spent at each purchase occasion.

Here we use these 3 simple characteristics to divide customers into actionable segments.

```{r calcRFM}

## Add columns for recency, frequency, and monetary value
recentDate <- max(purchases$date_of_purchase)
plotDat <- purchases %>%
    group_by(customer_id) %>%
    summarize(recency = min(recentDate - date_of_purchase),
              frequency = n(),
              monetary_value = mean(purchase_amount))

######################################
## Visually summarize these metrics ##
######################################
ggplot(data = plotDat, aes(x = frequency, y = monetary_value)) +
    geom_point(alpha = 0.1) +
    scale_y_log10(labels = scales::dollar) +
    annotation_logticks(sides = "l") +
    xlab("Purchase Frequency") +
    ylab("Average Purchase Amount") +
    getBaseTheme()


ggplot(data = plotDat, aes(x = recency / 365, y = monetary_value)) +
    geom_point(alpha = 0.1) +
    ## geom_hex() +
    scale_y_log10(labels = scales::dollar) +
    annotation_logticks(sides = "l") +
    xlab("Years Since Last Purchase (Recency)") +
    ylab("Average Purchase Amount") +
    getBaseTheme()


```

## Customer Segmentation

```{r segment}

## Note: The example code uses log of purchase amount because of the skewed distribution. This makes it so that larger purchase amounts are not as different as on a linear scale. You can experiment with using the commented line below to see the difference in clustering.
dat <- purchases %>%
    group_by(customer_id) %>%
    summarize(recency = as.numeric(min(recentDate - date_of_purchase)),
              monetary_value = mean(purchase_amount),
              ## monetary_value = mean(log10(purchase_amount)),
              frequency = n()) %>%
    as.data.frame()

## Put customer_id in the rownames of dat
rownames(dat) <- plotDat$customer_id
dat <- select(dat, -customer_id)

## Calculate distance matrix and hierarchical clustering
dis <- dist(scale(dat))
hc <- hclust(dis, method = "ward.D2")


## Plot the dendrogram
plot(hc)
abline(h = 50, col = "red")
cuts <- cutree(hc, h = 50)
segments <- paste0("Segment ", cuts, "\n(n = ", table(cuts)[cuts], ")")

```

## What does each segment look like?

We can make the same plot of Average Purchase Amount as a function of purchase frequency but color by segment.

```{r segmentViz1}

## Plot 2 variables at a time using cuts
## Shuffle row order to see overlapping points more clearly on plot
plotDat2 <- plotDat %>%
    mutate(segment = factor(segments)) %>%
    arrange(sample(nrow(plotDat)))

## Same plot as before but colored by cluster
ggplot(data = plotDat2,
       aes(x = frequency, y = monetary_value, color = segment)) +
    geom_point() +
    scale_y_log10(labels = scales::dollar) +
    annotation_logticks(sides = "l") +
    xlab("Purchase Frequency") +
    ylab("Average Purchase Amount") +
    getBaseTheme()

ggplot(data = plotDat2, aes(x = recency / 365, y = monetary_value, color = segment)) +
    geom_point() +
    scale_y_log10(labels = scales::dollar) +
    annotation_logticks(sides = "l") +
    xlab("Years Since Last Purchase (Recency)") +
    ylab("Average Purchase Amount") +
    getBaseTheme()

```

From the first plot we can see that all segments are clearly separated by these 2 variables except for 1 and 4. Segments 1 and 4 are separated by recency in the second plot.

We can visualize all 3 variables simultaneously in boxplots. This is a bit confusing because of the common scale when we are really showing 3 different things (frequency, dollars, years).

```{r segmentViz2, fig.height = 10}

## Plot all 3 variables in boxplots
plotDat3 <- plotDat2 %>%
    mutate(recency = (as.numeric(recency) %/% 365) + 1) %>% # integer division
    rename(`recency (years)` = recency) %>%
    gather(key = "RFM_type", value = "value",
           `recency (years)`, frequency, monetary_value)

ggplot(data = plotDat3, aes(x = RFM_type, y = value, color = segment)) +
    geom_boxplot(position = "dodge") +
    scale_y_log10(breaks = c(1, 10, 100, 1000)) +
    annotation_logticks(sides = "l") +
    facet_grid(segment ~ .) +
    getBaseTheme() +
    scale_color_discrete(guide = FALSE) +
    theme(strip.text = element_text(size = 12, face = "bold"))

## Summarize each segment by the median of each variable
plotDat %>%
    mutate(segment = cuts) %>%
    group_by(segment) %>%
    summarize(number = n(), median(frequency), median(recency), median(monetary_value)) %>%
    as.data.frame()

```

# Conclusions

Using hierarchical clustering we were able to define 6 clear customer segments with definining characterstics.

1. **Cold**
   - These customers shopped at our store 1-5 times over 4 years ago and haven't been back.
2. **Almost Regulars**
   - These customers spend a similar amount as the **Regulars** but don't buy as frequently (4 - 8 transactions)
3. **Regulars**
   - These customers are regulars.  They have shopped at least 8 times (some as many as 40 times). They have all made a purchase in the past year though their monetary value varies widely.
4. **Average Joes**
   - Customers who have shopped 1 or 2 times in the past 4 years and spend about $30 each time.
5. **High End**
   - A small segment of 281 customers who have shopped 1-10 times (the majority have shopped in the past year) and spend around $500.
6. **Money Items**
   - The smallest segment of 54 customers who don't shop frequently, but when they shop, they spend close to $2,000.
