---
title: "Foundations of Marketing Analytics: Module 0: Introduction"
author: "Stefan Avey"
date: "2017-01-02"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    df_print: kable
---


```{r setup, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(lubridate)
library(aveytoolkit)

options(stringsAsFactors = FALSE)
```

# Load the Data

```{r loadData}

## Load data
purchases <- read.delim("data/purchases.txt", header = FALSE)

## Add column names
colnames(purchases) <- c("customer_id", "purchase_amount", "date_of_purchase")

## Convert date and add column for purchase year
purchases <- purchases %>%
    mutate(date_of_purchase = ymd(date_of_purchase)) %>%
    mutate(year_of_purchase = year(date_of_purchase))

## Look at the data
head(purchases)
summary(purchases)

```

# Explore the Data

```{r summarizeData}

plotDat <- purchases %>%
    group_by(year_of_purchase) %>%
    summarize(counter = n(),
              avg_amount = mean(purchase_amount),
              sum_amount = sum(purchase_amount),
              med_amount = median(purchase_amount)) %>%
    ungroup() %>%
    arrange(year_of_purchase)

````

## How many purchases are made each year?

```{r purchasesPerYear}

ggplot(data = plotDat) +
    geom_bar(stat = "identity", aes(x = year_of_purchase, y = counter)) +
    scale_x_continuous(breaks = plotDat$year_of_purchase,
                       labels = plotDat$year_of_purchase) +
    xlab("Year") +
    ylab("Purchases") +
    getBaseTheme()

```

## How has the purchase amount changed over time?

```{r purchaseAmtPerYear}

ggplot(data = plotDat, aes(x = year_of_purchase, y = avg_amount)) +
    geom_point() +
    geom_line() +
    scale_x_continuous(breaks = plotDat$year_of_purchase,
                       labels = plotDat$year_of_purchase) +
    scale_y_continuous(labels = scales::dollar) +
    xlab("Year") +
    ylab("Average Purchase Amount") +
    getBaseTheme() +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank())

```

```{r medPurchaseAmtPerYear}

ggplot(data = plotDat, aes(x = year_of_purchase, y = med_amount)) +
    geom_point() +
    geom_line() +
    scale_x_continuous(breaks = plotDat$year_of_purchase,
                       labels = plotDat$year_of_purchase) +
    scale_y_continuous(labels = scales::dollar) +
    xlab("Year") +
    ylab("Median Purchase Amount") +
    getBaseTheme() +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank())

```

```{r purchaseAmtPerYear2}

lmod <- lm(data = plotDat %>% mutate(year = year_of_purchase - 2005),
           sum_amount ~ year)
ggplot(data = plotDat, aes(x = year_of_purchase, y = sum_amount)) +
    geom_smooth(method = lm, color = "#3182bd", se = FALSE) +
    geom_point() +
    geom_line() +
    annotate("text", x = 2008, y = 420000, color = "#3182bd",
             label = GetEqn(lmod), parse = TRUE) +
    scale_x_continuous(breaks = plotDat$year_of_purchase,
                       labels = plotDat$year_of_purchase) +
    scale_y_continuous(labels = scales::dollar) +
    xlab("Year") +
    ylab("Total Purchase Amount") +
    getBaseTheme() +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank()) +
    ggtitle("Linear Growth in Total Purchase Amount")

```

## What is the purchase distribution each year?

While looking at the number of purchases, average amount sold, and total amount sold are informative, it is helpful to explore the whole distribution of purchases per year.

```{r distributions}

ggplot(data = purchases, aes(x = year_of_purchase, y = purchase_amount)) +
    geom_violin(aes(group = year_of_purchase)) +
    scale_x_continuous(breaks = plotDat$year_of_purchase,
                       labels = plotDat$year_of_purchase) +
    scale_y_continuous(labels = scales::dollar) +
    xlab("Year") +
    ylab("Purchase Amount") +
    getBaseTheme() +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank())

```

# Conclusions

- From 2005--2015, the total purchase ammount has increased linearly at a rate of approximately $37,000 / year
- While the growth in total sales in 2005--2007 was due mainly to an increase in sales volume, the growth in 2012--2015 was due mainly to an increase in purchase amount.
